\systemEnding: Bootstrapping Domain-Specific Languages for Neurally-Guided Bayesian Program Learning

### 2018-06-14 archive.tex
more writing for archive posting

% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{icml2018} with \usepackage[nohyperref]{icml2018} above.
%\usepackage{hyperref}

@@

@@

\begin{table*} %[t!]
  \makebox[\textwidth][c]{
    \scriptsize
  \tabcolsep=4pt
  \renewcommand\code\texttt
  \renewcommand\codechar[1]{\texttt{"#1"}}
  \newcommand{\helpSize}{0.25cm}
  \begin{tabular}{>{\hspace{-0em}}c<{\hspace{-1em}}>{\hspace{-1em}}c<{\hspace{-1em}}>{\hspace{-2.5em}}c<{\hspace{-0.8em}}>{\hspace{-1em}}c<{\hspace{-1em}}}
    \toprule
    &{\normalsize List Functions}&{\normalsize Text Editing}&{\normalsize Symbolic Regression}\\\midrule
    \rotatebox[origin=c]{90}{\normalsize \pop{Programs} \& Tasks}&{\tabcolsep=7pt
      \begin{tabular}{cc}
        \begin{tabular}{c}
          \code{[7\, 2\, 3]}$\to $\code{[7\, 3]}         \\
          \code{[1\, 2\, 3\, 4]}$\to $\code{[3\, 4]} \\
          \code{[4\, 3\, 2\, 1]}$\to $\code{[4\, 3]} \\
          \pop{\code{$f(\ell) = $}\code{($f_1$ $\ell$ ($\lambda$ (x)}}\\
          \hspace{1.15cm}\pop{\code{(> x 2)))}}       \\
          \\
          \code{[2\, 7\, 8\, 1]}$\to $\code{8}               \\
          \hspace{0.15cm}\code{[3\, 19\, 14]}$\to $\code{19}                \\
          \pop{\code{$f(\ell) = $}\code{($f_2$ $\ell$)}}
        \end{tabular}
        &
        \hspace{-0.3cm}\begin{tabular}{c}
          \code{[7\, 3]}$\to $\code{False}                              \\
          \hspace{0.3cm}\code{[3]}$\to $\code{False}                    \\
          \hspace{-0.3cm}\code{[9\, 0\, 0]}$\to $\code{True\phantom{e}} \\
          \hspace{0.3cm}\code{[0]}$\to $\code{True\phantom{e}}                        \\
          \hspace{-0.3cm}\code{[0\, 7\, 3]}$\to $\code{True\phantom{e}}                \\
          \pop{\code{$f(\ell) = $}\code{($f_3$ $\ell$ 0)}}
        \end{tabular}
      \end{tabular}
    }
    &
    \hspace{-0.5cm}\begin{tabular}{c}
      +106 769-438$\to $106.769.438\\%&Nancy FreeHafer $\longrightarrow$ Dr. Nancy\\
      +83 973-831$\to $83.973.831\\
      \pop{$f(\text{\code{s}}) = $\code{(}$f_0$\code{  \codechar{.} \codechar{-}
      }}\\
      \hspace{1.25cm}\pop{\code{($f_0$ \codechar{.} \codechar{ }}}\\
      \hspace{1.4cm}\pop{\code{(cdr s)))}}\\
      ~\\
      Temple Anna H $\to $TAH\\
      Lara Gregori$\to $LG\\
      \pop{$f(\text{\code{s}}) = $\code{(}$f_2$\code{ s)}}\\
    \end{tabular}
    &
    \begin{tabular}{cc}
      \includegraphics[width = 3em]{figures/functions/4.png}&
      \includegraphics[width = 3em]{figures/functions/146}\\
      \pop{\code{$f($x$) = $($f_1$ x)}}&    \pop{\code{$f($x$) = $($f_6$ x)}}\\
      ~\\
      \includegraphics[width = 3em]{figures/functions/112.png}&
        \includegraphics[width = 3em]{figures/functions/92.png}
      \\
      \pop{\code{$f($x$) = $($f_4$ x)}}&    \pop{\code{$f($x$) = $($f_3$ x)}}\\

@@

The purpose of the DSL is to
offer a set of abstractions
that allow an agent to easily express solutions to the tasks at hand.
%In the \system algorithm we infer the DSL from a collection of frontiers.
Intuitively, we want the algorithm to
look at  the frontiers and
generalize beyond them, 
both so the DSL can better express the current solutions,
and  also so that the DSL might expose new abstractions
which will later be used to
discover more programs.
 Formally, we want the DSL maximizing $\int \lowerBound\;\mathrm{d}\theta$ (Sec.~\ref{mathematicalFraming}).
We replace this marginal with an AIC approximation, giving the following objective for DSL induction:
\begin{align}
\nonumber      \log \probability[\mathcal{D}]& + 
\argmax_{\theta}\Bigg(\log \probability[\theta|\mathcal{D}] - \|\theta\|_0 \\&+\sum_{x\in X}\log \sum_{p\in \mathcal{F}_x}\probability[x|p]\probability[p|\mathcal{D},\theta]\Bigg)
\label{AIC}
  \end{align}
We induce a DSL by searching locally through the space of DSLs,
proposing small changes to $\mathcal{D}$ until Eq.~\ref{AIC} fails to increase.
The search moves work by introducing new
$\lambda$-expressions into the DSL.
We propose these new expressions by extracting fragments of
programs already in the frontiers (Tbl.~\ref{fragmentExample}).
An important point here is that we are \emph{not} simply adding
subexpressions of programs to $\mathcal{D}$, as done in the EC algorithm~\cite{Dechter:2013:BLV:2540128.2540316} and other prior work~\cite{DBLP:conf/ecai/LinDETM14}.  Instead, we are
extracting fragments that unify with programs in the frontiers.  This
idea of storing and reusing fragments of expressions comes from
Fragment Grammars~\cite{tim} and Tree-Substitution
Grammars~\cite{cohn2010inducing}, and is closely related to the idea
of antiunification~\cite{henderson2013cumulative}.
\begin{figure*}[!ht]\tabcolsep=2pt
\centering  \begin{minipage}[c]{0.2\textwidth}
  \begin{tikzpicture}[scale=0.7]
    %% \node[rotate=30] at (-2,0) {\begin{tabular}{c}
    %%     \footnotesize Program:\\
    %%     \code{($\lambda$ (x) (+ (- x) 1))}
    %% \end{tabular}};
    %\node at (,0.5) {\code{cons}};
    \node [rotate=90] at (-2.3,-0.5) {\small program};

@@

%\subsubsection{List Processing}\label{listSection}
\textbf{List Processing:} Synthesizing programs that manipulate data structures is a widely studied
problem in the programming languages community~\cite{feser2015synthesizing}.
%with applications to computer aided programming~\cite{solar2008program}.
We consider this problem within the context of learning functions that
manipulate lists, and also perform arithmetic operations upon lists (Table~\ref{listExamples}).
We created 236 human-interpretable list manipulation tasks, each with 15
input/output examples.
%% Our data set is interesting in three major ways: many of the tasks
%% require complex solutions; the tasks were not generated from some
%% latent DSL, and the agent must learn to solve these complicated
%% problems from only 236 tasks.
%% Our data set assumes arithmetic operations as well as sequence operations,
%% so we additionally provide our system with the following arithmetic
%% primitives: \code{mod}, \code{*}, \code{>}, \code{is-square},
%% \code{is-prime}.
%% evaluating  on a 50/50 test/train split.
In solving these tasks, the system
composed 38 new subroutines, and rediscovered the higher-order
function \code{filter} ($f_1$ in Table~\ref{initialExampleDSL}, left).%% , yielding a more expressive DSL more closely
%% matching the domain (Tbl.~\ref{initialExampleDSL}, left).
\begin{figure}[b]\centering
\vspace{-0.5cm}  \begin{tabular}{lll}
    \toprule
    Name & Input & Output \\\midrule
    repeat-3 & [7\, 0] & [7\, 0\, 7\, 0\, 7\, 0] \\
%    drop-3 & [0\, 3\, 8\, 6\, 4] & [6\, 4] \\
    rotate-2 & [8\, 14\, 1\, 9] & [1\, 9\, 8\, 14] \\
%    count-head-in-tail & [1\, 2\, 1\, 1\, 3] & 2 \\
    keep-div-5 & [5\, 9\, 14\, 6\, 3\, 0] & [5\, 0] \\
    product & [7\, 1\, 6\, 2] & 84 \\
    \bottomrule
  \end{tabular}
  \captionof{table}{Some tasks in our list function domain.}\label{listExamples}\vspace{-0.5cm}
\end{figure}

@@

\subsection{Symbolic Regression: Programs from visual input}\label{regressionSection}
We apply \system
to symbolic regression problems.  Here, the
agent observes points along the curve of a function, and must write a
program that fits those points.  We initially equip our learner with
addition, multiplication, and division, and task it with solving
100 %% symbolic regression
problems, each either a polynomial or rational function.  The recognition model is a
convnet that observes an image of the target function's
graph (Fig.~\ref{functions}) --- visually, different kinds of
polynomials and rational functions produce different kinds of graphs,
and so the convnet can  look at a graph and predict
what kind of function best explains it.
These programs can contain real numbers,
which we set with gradient descent,
and penalize continuous parameters by
incorporating a BIC penalty into the likelihood model $\probability[x|p]$.
%% A key difficulty, however, is
%% that these problems are best solved with programs containing real
%% numbers.  Our solution to this difficulty is to enumerate
%%  programs with real-valued parameters, and then fit those
%% parameters by automatically differentiating through the programs the
%% system writes and use gradient descent to fit the parameters.
%% We define the likelihood model, $\probability[x|p]$, by assuming a Gaussian noise model for the input/output examples,
%% and penalize the use of real-valued parameters using the BIC~\cite{Bishop:2006:PRM:1162264}.
\begin{figure}\vspace{-0.0cm} \newcommand{\functionSize}{1.1cm}\centering
  \begin{minipage}[c]{0.3\columnwidth}%\vspace{-0.3cm}
  \includegraphics[width = \functionSize]{figures/functions/6.png}
  \includegraphics[width = \functionSize]{figures/functions/48.png}\\
  \includegraphics[width = \functionSize]{figures/functions/102.png}
  \includegraphics[width = \functionSize]{figures/functions/116.png}\\
  \vspace{1pt}%\includegraphics[width = \functionSize]{figures/functions/181.png}
  %\includegraphics[width = \functionSize]{figures/functions/160.png}
  \includegraphics[width = \functionSize]{figures/functions/160.png}
    \includegraphics[width = \functionSize]{figures/functions/149.png}
  \end{minipage}
  \begin{minipage}[c]{0.69\columnwidth}    
    \caption{Recognition model input for symbolic regression. DSL learns subroutines for polynomials (top rows) and rational functions (bottom) while the recognition  model jointly learns to look at a graph of the function (left) and predict which learned subroutines best explain the observation.}\label{functions}\vspace{-1cm}
        \end{minipage}
\end{figure}
We learn a DSL containing 13 new functions,
mainly templates for polynomials of different orders or ratios of polynomials.
The model also learns to find programs minimizing the number of continuous parameters ---
learning to represent linear functions with 
\code{(* real (+ x real))}, which has two continuous parameters, and represents quartic functions using $f_4$ in the rightmost column of Fig.~\ref{initialExampleDSL}
which has five continuous parameters.
This phenomenon arises from our Bayesian framing.

\subsection{Learning from Scratch}
A long-standing dream within the program induction community
is ``learning from scratch'': starting with a \emph{minimal} Turing-complete programming language,
and then learning to solve a wide swath of
induction problems~\cite{solomonoff1964formal}.
All existing systems,
including ours,
fall far short of this dream,
%and it is unclear (and we believe unlikely)
and we believe it is unlikely
that this dream could ever be fully realized.
%How far can we push in this direction?
``Learning from scratch'' is subjective, but a reasonable
starting point is the set of primitives provided in 1959
Lisp~\cite{mccarthy1960recursive}: these include
conditionals, recursion, arithmetic, and the 
list operators \code{cons}, \code{car}, \code{cdr}, and \code{nil}.
A  basic first goal is to start with
these primitives,
and then recover a DSL that
more closely resembles modern functional languages like Haskell and OCaml.
Recall (Sec.~\ref{sequences})
that we initially provided our system with functional programming routines like
\code{map} and \code{fold}.

We ran the following experiment: \system was given a subset of
the 1959 Lisp primitives, and tasked with solving 22
programming exercises.  After running for 93 hours on 64 CPUs, our
algorithm solves these exercises, along the way assembling a DSL
with a modern repertoire of
functional programming idioms and subroutines, including \code{map},
\code{fold}, \code{zip}, \code{unfold}, \code{index}, \code{length},
and  arithmetic operations like 
building lists of natural numbers between an interval.
%% We did not use the recognition model for this experiment:
%% a bottom-up pattern recognizer is of little use
%% for acquiring this abstract knowledge
%% from less than two dozen problems.

>>>>

\usepackage[section]{placeins}
% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{icml2018} with \usepackage[nohyperref]{icml2018} above.
%\usepackage{hyperref}

@@

\newcommand{\greenCode}[1]{{\footnotesize\popp{\code{#1}}}}

@@

\begin{table*}%[t!]
  \makebox[\textwidth][c]{
    \scriptsize
  \tabcolsep=4pt
  \renewcommand\code\texttt
  \renewcommand\codechar[1]{\texttt{"#1"}}
  \newcommand{\helpSize}{0.25cm}
  \begin{tabular}{>{\hspace{-0em}}c<{\hspace{-1em}}>{\hspace{-1em}}c<{\hspace{-1em}}>{\hspace{-2.5em}}c<{\hspace{-0.8em}}>{\hspace{-1em}}c<{\hspace{-1em}}}
    \toprule
    &{\normalsize List Functions}&{\normalsize Text Editing}&{\normalsize Symbolic Regression}\\\midrule
    \rotatebox[origin=c]{90}{\normalsize \pop{Programs} \& Tasks}&{\tabcolsep=7pt
      \begin{tabular}{cc}
        \begin{tabular}{c}
          \code{[7\, 2\, 3]}$\to $\code{[7\, 3]}         \\
          \code{[1\, 2\, 3\, 4]}$\to $\code{[3\, 4]} \\
          \code{[4\, 3\, 2\, 1]}$\to $\code{[4\, 3]} \\
          \pop{\code{$f(\ell) = $}\code{($f_1$ $\ell$ ($\lambda$ (x)}}\\
          \hspace{1.15cm}\pop{\code{(> x 2)))}}       \\
          \\
          \code{[2\, 7\, 8\, 1]}$\to $\code{8}               \\
          \hspace{0.15cm}\code{[3\, 19\, 14]}$\to $\code{19}                \\
          \pop{\code{$f(\ell) = $}\code{($f_2$ $\ell$)}}
        \end{tabular}
        &
        \hspace{-0.3cm}\begin{tabular}{c}
          \code{[7\, 3]}$\to $\code{False}                              \\
          \hspace{0.3cm}\code{[3]}$\to $\code{False}                    \\
          \hspace{-0.3cm}\code{[9\, 0\, 0]}$\to $\code{True\phantom{e}} \\
          \hspace{0.3cm}\code{[0]}$\to $\code{True\phantom{e}}                        \\
          \hspace{-0.3cm}\code{[0\, 7\, 3]}$\to $\code{True\phantom{e}}                \\
          \pop{\code{$f(\ell) = $}\code{($f_3$ $\ell$ 0)}}
        \end{tabular}
      \end{tabular}
    }
    &
    \hspace{-0.5cm}\begin{tabular}{c}
      +106 769-438$\to $106.769.438\\%&Nancy FreeHafer $\longrightarrow$ Dr. Nancy\\
      +83 973-831$\to $83.973.831\\
      \pop{$f(\text{\code{s}}) = $\code{(}$f_0$\code{  \codechar{.} \codechar{-}
      }}\\
      \hspace{1.25cm}\pop{\code{($f_0$ \codechar{.} \codechar{ }}}\\
      \hspace{1.4cm}\pop{\code{(cdr s)))}}\\
      ~\\
      Temple Anna H $\to $TAH\\
      Lara Gregori$\to $LG\\
      \pop{$f(\text{\code{s}}) = $\code{(}$f_2$\code{ s)}}\\
    \end{tabular}
    &
    \begin{tabular}{cc}
      \includegraphics[width = 3em]{figures/functions/4.png}&
      \includegraphics[width = 3em]{figures/functions/146}\\
      \pop{\code{$f($x$) = $($f_1$ x)}}&    \pop{\code{$f($x$) = $($f_6$ x)}}\\
      ~\\
      \includegraphics[width = 3em]{figures/functions/112.png}&
        \includegraphics[width = 3em]{figures/functions/92.png}
      \\
      \pop{\code{$f($x$) = $($f_4$ x)}}&    \pop{\code{$f($x$) = $($f_3$ x)}}\\

@@

The purpose of the DSL is to
offer a set of abstractions
that allow an agent to easily express solutions to the tasks at hand.
%In the \system algorithm we infer the DSL from a collection of frontiers.
Intuitively, we want the algorithm to
look at  the frontiers and
generalize beyond them, 
both so the DSL can better express the current solutions,
and  also so that the DSL might expose new abstractions
which will later be used to
discover more programs.
 Formally, we want the DSL maximizing $\int \lowerBound\;\mathrm{d}\theta$ (Sec.~\ref{mathematicalFraming}).
We replace this marginal with an AIC approximation, giving the following objective for DSL induction:
\begin{align}
\nonumber      \log \probability[\mathcal{D}]& + 
\argmax_{\theta}\Bigg(\log \probability[\theta|\mathcal{D}] - \|\theta\|_0 \\&+\sum_{x\in X}\log \sum_{p\in \mathcal{F}_x}\probability[x|p]\probability[p|\mathcal{D},\theta]\Bigg)
\label{AIC}
  \end{align}
We induce a DSL by searching locally through the space of DSLs,
proposing small changes to $\mathcal{D}$ until Eq.~\ref{AIC} fails to increase.
The search moves work by introducing new
$\lambda$-expressions into the DSL.
We propose these new expressions by extracting fragments of
programs already in the frontiers (Tbl.~\ref{fragmentExample}).
An important point here is that we are \emph{not} simply adding
subexpressions of programs to $\mathcal{D}$, as done in the EC algorithm~\cite{Dechter:2013:BLV:2540128.2540316} and other prior work~\cite{DBLP:conf/ecai/LinDETM14}.  Instead, we are
extracting fragments that unify with programs in the frontiers.  This
idea of storing and reusing fragments of expressions comes from
Fragment Grammars~\cite{tim} and Tree-Substitution
Grammars~\cite{cohn2010inducing}, and is closely related to the idea
of antiunification~\cite{henderson2013cumulative}.

%\vfill\null

%\columnbreak

\begin{figure*}[!htb]\tabcolsep=2pt
\centering  \begin{minipage}[c]{0.2\textwidth}
  \begin{tikzpicture}[scale=0.7]
    %% \node[rotate=30] at (-2,0) {\begin{tabular}{c}
    %%     \footnotesize Program:\\
    %%     \code{($\lambda$ (x) (+ (- x) 1))}
    %% \end{tabular}};
    %\node at (,0.5) {\code{cons}};
    \node [rotate=90] at (-2.3,-0.5) {\small program};

@@

%\subsubsection{List Processing}\label{listSection}
\textbf{List Processing:} Synthesizing programs that manipulate data structures is a widely studied
problem in the programming languages community~\cite{feser2015synthesizing}.
%with applications to computer aided programming~\cite{solar2008program}.
We consider this problem within the context of learning functions that
manipulate lists, and also perform arithmetic operations upon lists (Table~\ref{listExamples}).
We created 236 human-interpretable list manipulation tasks, each with 15
input/output examples.
%% Our data set is interesting in three major ways: many of the tasks
%% require complex solutions; the tasks were not generated from some
%% latent DSL, and the agent must learn to solve these complicated
%% problems from only 236 tasks.
%% Our data set assumes arithmetic operations as well as sequence operations,
%% so we additionally provide our system with the following arithmetic
%% primitives: \code{mod}, \code{*}, \code{>}, \code{is-square},
%% \code{is-prime}.
%% evaluating  on a 50/50 test/train split.
In solving these tasks, the system
composed 38 new subroutines, and rediscovered the higher-order
function \code{filter} ($f_1$ in Table~\ref{initialExampleDSL}, left).%% , yielding a more expressive DSL more closely
%% matching the domain (Tbl.~\ref{initialExampleDSL}, left).
\begin{figure}[b]\centering
\vspace{-0.5cm}  \begin{tabular}{lll}
    \toprule
    Name & Input & Output \\\midrule
    repeat-3 & [7\, 0] & [7\, 0\, 7\, 0\, 7\, 0] \\
    drop-3 & [0\, 3\, 8\, 6\, 4] & [6\, 4] \\
    rotate-2 & [8\, 14\, 1\, 9] & [1\, 9\, 8\, 14] \\
    count-head-in-tail & [1\, 2\, 1\, 1\, 3] & 2 \\
    keep-div-5 & [5\, 9\, 14\, 6\, 3\, 0] & [5\, 0] \\
    product & [7\, 1\, 6\, 2] & 84 \\
    \bottomrule
  \end{tabular}
  \captionof{table}{Some tasks in our list function domain.}\label{listExamples}\vspace{-0.5cm}
\end{figure}

@@

\subsection{Symbolic Regression: Programs from visual input}\label{regressionSection}
We apply \system
to symbolic regression problems.  Here, the
agent observes points along the curve of a function, and must write a
program that fits those points.  We initially equip our learner with
addition, multiplication, and division, and task it with solving
100 %% symbolic regression
problems, each either a polynomial or rational function.  The recognition model is a
convnet that observes an image of the target function's
graph (Fig.~\ref{functions}) --- visually, different kinds of
polynomials and rational functions produce different kinds of graphs,
and so the convnet can  look at a graph and predict
what kind of function best explains it.
A key difficulty, however, is
that these problems are best solved with programs containing real
numbers.  Our solution to this difficulty is to enumerate
 programs with real-valued parameters, and then fit those
parameters by automatically differentiating through the programs the
system writes and use gradient descent to fit the parameters.
We define the likelihood model, $\probability[x|p]$, by assuming a Gaussian noise model for the input/output examples,
and penalize the use of real-valued parameters using the BIC~\cite{Bishop:2006:PRM:1162264}.
\begin{figure}\vspace{-0.0cm} \newcommand{\functionSize}{1.1cm}\centering
  \begin{minipage}[c]{0.3\columnwidth}%\vspace{-0.3cm}
  \includegraphics[width = \functionSize]{figures/functions/6.png}
  \includegraphics[width = \functionSize]{figures/functions/48.png}\\
  \includegraphics[width = \functionSize]{figures/functions/102.png}
  \includegraphics[width = \functionSize]{figures/functions/116.png}\\
  \vspace{1pt}%\includegraphics[width = \functionSize]{figures/functions/181.png}
  %\includegraphics[width = \functionSize]{figures/functions/160.png}
  \includegraphics[width = \functionSize]{figures/functions/160.png}
    \includegraphics[width = \functionSize]{figures/functions/149.png}
  \end{minipage}
  \begin{minipage}[c]{0.69\columnwidth}    
    \caption{Recognition model input for symbolic regression. DSL learns subroutines for polynomials (top rows) and rational functions (bottom) while the recognition  model jointly learns to look at a graph of the function (left) and predict which learned subroutines best explain the observation.}\label{functions}\vspace{-1cm}
        \end{minipage}
\end{figure}
We learn a DSL containing 13 new functions,
mainly templates for polynomials of different orders or ratios of polynomials.
The model also learns to find programs minimizing the number of continuous parameters ---
learning to represent linear functions with 
\code{(* real (+ x real))}, which has two continuous parameters, and represents quartic functions using $f_4$ in the rightmost column of Fig.~\ref{initialExampleDSL}
which has five continuous parameters.
This phenomenon arises from our Bayesian framing:
both the generative model's bias toward shorter programs,
and the likelihood model's BIC penalty.

\subsection{Learning from Scratch}
A long-standing dream within the program induction community
is ``learning from scratch'': starting with a \emph{minimal} Turing-complete programming language,
and then learning to solve a wide swath of
induction problems~\cite{solomonoff1964formal}.
All existing systems,
including ours,
fall far short of this dream,
and it is unclear (and we believe unlikely)
that this dream could ever be fully realized.
How far can we push in this direction?
``Learning from scratch'' is subjective, but a reasonable
starting point is the set of primitives provided in 1959
Lisp~\cite{mccarthy1960recursive}: these include
conditionals, recursion, arithmetic, and the 
list operators \code{cons}, \code{car}, \code{cdr}, and \code{nil}.
A  basic first goal is to start with
these primitives,
and then recover a DSL that
more closely resembles modern functional languages like Haskell and OCaml.
Recall (Sec.~\ref{sequences})
that we initially provided our system with functional programming routines like
\code{map} and \code{fold}.

We ran the following experiment: \system was given a subset of
the 1959 Lisp primitives, and tasked with solving 22
programming exercises. A key difference between
this setup and our previous experiments is that,
for this experiment,
the system is
given primitive recursion,
whereas previously we had sequestered recursion within
higher-order functions like \code{map}, \code{fold}, and \code{unfold}.

After running for 93 hours on 64 CPUs, our
algorithm solves these 22 exercises, along the way assembling a DSL
with a modern repertoire of
functional programming idioms and subroutines, including \code{map},
\code{fold}, \code{zip}, \code{unfold}, \code{index}, \code{length},
and  arithmetic operations like 
building lists of natural numbers between an interval (see Figure~\ref{learningFromScratch}).
We did not use the recognition model for this experiment:
a bottom-up pattern recognizer is of little use
for acquiring this abstract knowledge
from less than two dozen problems.

\begin{figure}  \newcommand{\helpSize}{0.25cm}
  \begin{tabular}{cc}\toprule
    \rotatebox[origin=c]{90}{\normalsize \pop{Programs} \& Tasks}\\\midrule
    \rotatebox[origin=c]{90}{\normalsize \popp{DSL}}&
    \begin{tabular}{l}
      \greenCode{$f_0($x$)\,=\,$(+ x 1)}\\
      \hspace{\helpSize}($f_0$: \emph{increment})\\
      \greenCode{$f_1($x$)\,=\,$(- x 1)}\\
      \hspace{\helpSize}($f_1$: \emph{decrement})\\
        \popp{\code{$f_2($p$,$f$,$n$,$x$)\,=\,$(if (p x) nil}}\\
      \phantom{\code{$f_1($f$,$l$,$x$)\,=\,$(if }}}\popp{\code{(cons (f x) ($f_2$ (n x))))}}\\
        \hspace{\helpSize}($f_2$: \emph{unfold})\\
        \popp{\code{$f_3($i$,$l$)\,=\,$(if (= i 0) (car l)}}\\
      \phantom{\code{$f_1($f$,$l$,$x$)\,=\,$(if }}}\popp{\code{($f_3$ ($f_1$ i) (cdr l))))}}\\
        \hspace{\helpSize}($f_3$: \emph{index})\\
        \popp{\code{$f_4($f$,$l$,$x$)\,=\,$(if (empty? l) x}}\\
      \phantom{\code{$f_4($f$,$l$,$x$)\,=\,$(if }}}\popp{\code{(f (car l) ($f_4$ (cdr l))))}}\\
        \hspace{\helpSize}($f_4$: \emph{fold})\\

      \end{tabular}
    %% : (lambda (#(#(lambda (lambda (lambda (lambda (fix1 $0 (lambda (lambda (if (empty? $0) $3 ($4 ($5 $0) ($1 (cdr $0))))))))))) (lambda (car $0))) (lambda (lambda (+ $0 $1))) 0 $0))
    %% (lambda (lambda (lambda (lambda (fix1 $0 (lambda (lambda (#(lambda (lambda (lambda (if $0 empty (cons $1 $2))))) ($1 ($3 $0)) ($4 $0) ($5 $0)))))))))
    %% (lambda (lambda (fix1 $0 (lambda (lambda (#(lambda (lambda (lambda (if $0 empty (cons $1 $2))))) (#(lambda (lambda (fix1 $0 (lambda (lambda (if (empty? $0) empty (cons ($3 (car $0)) ($1 (cdr $0))))))))) (lambda (#(+ 1) $0)) ($1 (#(lambda (- $0 1)) $0))) 0 (eq? $0 $3)))))))
    \\\bottomrule 
    \end{tabular}
  \caption{Bootstrapping a standard library of functional programming routines, starting from recursion along with primitive operations found in 1959 Lisp.}\label{learningFromScratch}
  \end{figure}
